STA4173 – Final Project
Diabetes Health Indicators Analysis
By Jack Griffin

Goals of this analysis:
1) Explore diabetes prevalence and key lifestyle/demographic risk factors
2) Build a multinomial logistic regression model (3-category outcome)
3) Build a binary logistic regression model (collapsed outcome)
4) Compare model performance and interpretability

# STEP ONE: Setup -- Packages & Libraries
```{r}
# Run this install line ONCE if needed (uncomment it)
# install.packages(c("tidyverse","nnet","car","caret","janitor","gridExtra","corrplot","pscl","pROC"))

library(tidyverse)   # ggplot2, dplyr, etc.
library(nnet)        # multinomial logistic regression
library(car)         # VIF
library(caret)       # confusion matrix / model metrics
library(janitor)     # frequency tables
library(gridExtra)   # arranging multiple plots
library(corrplot)    # correlation heatmap
library(pscl)        # pseudo R-squared
library(pROC)        # ROC curves
```

# STEP TWO: Load & Check Dataset
```{r}
# Load CSV dataset (make sure it's in your working directory)
# df <- read.csv("diabetes_012_health_indicators_BRFSS2015.csv")
# Uncomment the above line to load in data. I was having issues so I am using file.choose.

df <- read.csv(file.choose())


# Convert outcome variable to labeled factor
df$Diabetes_012 <- factor(df$Diabetes_012,
                          levels = c(0, 1, 2),
                          labels = c("NoDiabetes", "Prediabetes", "Diabetes"))

# Basic structure checks
dim(df)         # number of rows and columns
names(df)       # variable names
glimpse(df)     # types and preview
summary(df)     # quick descriptive stats

# Total missing values
sum(is.na(df))
```

# STEP THREE: Explore Continuous / Ordinal Variables (Summary Statistics)
```{r}
# The only variables appropriate for numerical summary stats in the dataset: BMI, Age, GenHlth, PhysHlth, MentHlth, Education, Income.

# Select continuous and ordinal variables
cont_vars <- c("BMI","Age","GenHlth","PhysHlth",
               "MentHlth","Education","Income")

# Generate summary statistics
df %>%
  select(all_of(cont_vars)) %>%
  summary()
```

# STEP FOUR: Explore Binary Variables (Frequency Tables)
```{r}
binary_vars <- c("HighBP","HighChol","CholCheck","Smoker",
                 "Stroke","HeartDiseaseorAttack","PhysActivity",
                 "Fruits","Veggies","HvyAlcoholConsump",
                 "AnyHealthcare","NoDocbcCost","DiffWalk","Sex")

# Frequency tables for each binary predictor
freq_tables <- lapply(binary_vars, function(v){
  janitor::tabyl(df[[v]]) %>% 
    janitor::adorn_percentages("col") %>% 
    janitor::adorn_pct_formatting(digits = 1)
})

names(freq_tables) <- binary_vars
freq_tables
```

# STEP FIVE: Response Variable Distribution (Diabetes Status)
```{r}
df %>% 
  count(Diabetes_012) %>% 
  janitor::adorn_totals("row")
```

# STEP SIX: Histograms for Key Continuous Variables
```{r}
# BMI distribution
ggplot(df, aes(x = BMI)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(title = "BMI Distribution", x = "BMI", y = "Count")

# Physical health days distribution
ggplot(df, aes(x = PhysHlth)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(title = "Physical Health Days Distribution", x = "PhysHlth", y = "Count")

# Mental health days distribution
ggplot(df, aes(x = MentHlth)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(title = "Mental Health Days Distribution", x = "MentHlth", y = "Count")
```

# STEP SEVEN: Bar Plots for Ordinal Predictors
```{r}
# Education distribution
ggplot(df, aes(x = factor(Education))) +
  geom_bar(fill = "darkorange") +
  labs(title = "Education Level Distribution", x = "Education Level", y = "Count")

# Income distribution
ggplot(df, aes(x = factor(Income))) +
  geom_bar(fill = "darkorange") +
  labs(title = "Income Level Distribution", x = "Income Level", y = "Count")

# General Health distribution
ggplot(df, aes(x = factor(GenHlth))) +
  geom_bar(fill = "darkorange") +
  labs(title = "General Health Rating (1=Excellent, 5=Poor)",
       x = "GenHlth Category", y = "Count")
```

# STEP EIGHT: Bar Plots for Binary Predictors
```{r}
# Smoking status
ggplot(df, aes(x = factor(Smoker))) +
  geom_bar(fill = "purple") +
  labs(title = "Smoking Status", x = "Smoker (0/1)", y = "Count")

# Physical activity
ggplot(df, aes(x = factor(PhysActivity))) +
  geom_bar(fill = "purple") +
  labs(title = "Physical Activity", x = "PhysActivity (0/1)", y = "Count")

# High blood pressure
ggplot(df, aes(x = factor(HighBP))) +
  geom_bar(fill = "purple") +
  labs(title = "High Blood Pressure", x = "HighBP (0/1)", y = "Count")
```

# STEP NINE: Correlation Heatmap for Numeric Variables
```{r}
numeric_vars <- df %>%
  select(BMI, Age, HighBP, HighChol, GenHlth, PhysHlth, MentHlth)

corr_matrix <- cor(numeric_vars)

corrplot(corr_matrix, method = "color",
         addCoef.col = "black", tl.col = "black")
```

# STEP TEN: Train/Test Split for Model Building
```{r}
set.seed(123)   # for reproducibility

# Create training index
train_index <- createDataPartition(df$Diabetes_012, p = 0.7, list = FALSE)

# Split data
train <- df[train_index, ]
test  <- df[-train_index, ]

dim(train)
dim(test)
```

# STEP ELEVEN: Rebalance training data to fix class imbalance
```{r}
train_bal <- upSample(
  x = train %>% select(-Diabetes_012),   # predictors only
  y = train$Diabetes_012                 # outcome
)

# Renames the outcome column to "Class"
train_bal$Class <- factor(train_bal$Class,
                          levels = c("NoDiabetes","Prediabetes","Diabetes"))
```

# STEP TWELVE: Fit the Multinomial Logistic Regression Model
```{r}
# Fit multinomial logistic regression
model_multi <- multinom(
  Class ~ BMI + Age + PhysActivity + Smoker +
    HighBP + HighChol + GenHlth + PhysHlth + MentHlth +
    DiffWalk + Income + Education,
  data = train_bal,
  trace = FALSE)

summary(model_multi)
```

# STEP THIRTEEN: Convert Coefficients to Odd Ratios
```{r}
# Extract coefficients
coef_multi <- coef(model_multi)

# Convert to odds ratios
OR_multi <- exp(coef_multi)

OR_multi
```

# STEP FOURTEEN: Model Predictions on Test Set
```{r}
# Predict class probabilities
pred_probs <- predict(model_multi, newdata = test, type = "probs")

# Predict class labels
pred_class <- predict(model_multi, newdata = test)
```

# STEP FIFTEEN: Confusion Matrix & Accuracy
```{r}
conf_matrix <- confusionMatrix(as.factor(pred_class),
                               as.factor(test$Diabetes_012))

conf_matrix
```

# STEP SIXTEEN: Pseudo R-squared (Model Fit)
```{r}
# Fit intercept-only null model
null_model <- multinom(Class ~ 1, data = train_bal, trace = FALSE)

# Extract deviances
ll_null <- null_model$deviance
ll_full <- model_multi$deviance

# McFadden's Pseudo R²
pseudo_R2 <- 1 - (ll_full / ll_null)
pseudo_R2
```

# STEP SEVENTEEN: Check Multicollinearity (VIF)
```{r}
vif_values <- car::vif(model_multi)
vif_values
```

# STEP EIGHTEEN: Create the Binary Outcome Variable for Binary Model
```{r}
# Create binary outcome: 0 = NoDiabetes, 1 = Prediabetes/Diabetes
df$Diabetes_binary <- ifelse(df$Diabetes_012 == "NoDiabetes", 0, 1)
df$Diabetes_binary <- factor(df$Diabetes_binary)
```

# STEP NINETEEN: Train/Test Split for Binary Model
```{r}
set.seed(123)

train_index_bin <- createDataPartition(df$Diabetes_binary, p = 0.7, list = FALSE)

train_bin <- df[train_index_bin, ]
test_bin  <- df[-train_index_bin, ]

dim(train_bin)
dim(test_bin)
```

# STEP TWENTY: Fit Binary Logistic Regression Model
```{r}
model_bin <- glm(
  Diabetes_binary ~ BMI + Age + PhysActivity + Smoker +
    HighBP + HighChol + GenHlth + PhysHlth + MentHlth +
    DiffWalk + Income + Education,
  data = train_bin,
  family = binomial)

summary(model_bin)
```

# STEP TWENTY ONE: Convert Coefficients to Odds Ratios (Binary Model)
```{r}
coef_bin <- coef(model_bin)
OR_bin <- exp(coef_bin)

OR_bin
```

# STEP TWENTY TWO: Binary Model Predictions on Test Set
```{r}
# Predict probabilities and class labels
pred_probs_bin <- predict(model_bin, newdata = test_bin, type = "response")
pred_class_bin <- ifelse(pred_probs_bin > 0.5, 1, 0)

pred_class_bin <- factor(pred_class_bin, levels = c(0,1))
```

# STEP TWENTY THREE: Confusion Matrix & Accuracy
```{r}
conf_matrix_bin <- confusionMatrix(
  pred_class_bin,
  test_bin$Diabetes_binary)

conf_matrix_bin
```

# STEP TWENTY FOUR: ROC Curve & AUC
```{r}
# ROC curve
roc_obj <- roc(test_bin$Diabetes_binary, pred_probs_bin)
plot(roc_obj, col = "blue", main = "ROC Curve for Binary Logistic Regression")

# Pulling AUC value
auc_value <- auc(roc_obj)
auc_value
```

# STEP TWENTY FIVE: Psuedo R-squared for Binary Model
```{r}
ll_null_bin <- glm(Diabetes_binary ~ 1, data = train_bin, family = binomial)$deviance
ll_full_bin <- model_bin$deviance

pseudo_R2_bin <- 1 - (ll_full_bin / ll_null_bin)
pseudo_R2_bin
```